---
title: "Assignment 4: Water Quality in Rivers"
author: "Lindsay Roth"
output: html_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Hydrologic Data Analysis on water quality in rivers.

## Directions
1. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single HTML file.
8. After Knitting, submit the completed exercise (HTML file) to the dropbox in Sakai. Add your last name into the file name (e.g., "A04_Chamberlin.html") prior to submission.

The completed exercise is due on 25 September 2019 at 9:00 am.

## Setup

1. Verify your working directory is set to the R project file, 
2. Load the tidyverse, dataRetrieval, cowplot, xts and dygraphs packages. 
3. Set your ggplot theme (can be theme_classic or something else)

```{r setup}
getwd()



packages <- c("tidyverse", 
              "dataRetrieval", 
              "cowplot", 
              "xts",
              "dygraphs", 
              "lubridate",
              "base")
invisible(lapply(packages, library, character.only = TRUE)) 


theme_set(theme_classic(base_size = 12))
```

## Hypoxia

This assignment will look at another measure of water quality - oxygen concentration. Though not directly important for human health, oxygen in the water column is very important for aquatic life, and so is considered a measure of water quality. Hypoxia (low oxygen) has many different definitions. For this assignment, we will use 2 mg/L O~2~ as our cut-off.


4. Import the oxygen water quality data from New Hope Creek at Blands (using `readNWISqw()`, site code `02097314`, parameter code `00300`). Make a data frame called `O2.dat` that includes only the Date and O~2~ concentration values. Give your data frame understandable column names.

```{r read_data}
data.raw <- readNWISqw(siteNumbers = "02097314", #New Hope Creek at Blands
                     parameterCd = "00300", #oxygen water quality data
                     startDate = "",
                     endDate = "")

dat.02 <- data.raw %>%
  select(c(sample_dt, result_va))

class(dat.02$Date)
class(dat.02$Oxygen)

names(dat.02) <- c("Date","Oxygen")
```

5. Create a ggplot of oxygen concentrations over time. Include a horizontal line at 2 mg/l to show the hypoxia cutoff. 

```{r}
Oxygen.Plot <- ggplot(dat.02) +
  geom_point(aes(x = Date, y = Oxygen)) +
  geom_hline(yintercept = 2, color = "blue") +
  labs(y = "Oxygen (mg/L)") +
  ggtitle("Oxygen Concentrations Over Time")
  

print(Oxygen.Plot)
```

6. What do you notice about the frequency of hypoxia overtime?

> Hypoxic conditions have stopped after the year 1984, but oxygen levels appear to be steadily decreasing over time, even if conditions are not hypoxic.

## Nutrients

7. Often times hypoxia is associated with high nutrient concentrations, because abundant nutrients promote biomass growth which increases respiration and depletes oxygen concentrations in the water (remember how oxygen concentrations were very low in the hypolimnion from the Physical Properties of Lakes week). Create a new data frame, called `nutrients.dat` with total nitrogen (parameter code `00600`) and total phosphorus (parameter code `00665`) data from the USGS. Your data frame should have 3 columns, `Date`, `TotalNitrogen_mgl-N`, and `TotalPhosphorus_mgl-P`. 


```{r read_nutrients}
nutrients.dat <- readNWISqw(siteNumbers = "02097314", #New Hope Creek at Blands
                     parameterCd = c("00600",  #total Nitrogen
                                     "00665"), #total phosphorus
                     startDate = "",
                     endDate = "")

nutrients.skinny <- nutrients.dat %>%
  select(sample_dt,parm_cd,result_va) %>%
  group_by(sample_dt) %>%
  tibble::rowid_to_column() %>%
  spread(key = parm_cd, value = result_va) %>%
  select(-rowid,)
                     
names(nutrients.skinny) <- c("Date", "TotalNitrogen_mglN", "TotalPhosphorus_mglP")
```

8. Create two ggplots stacked with cowplot that show nutrient concentrations over time.

```{r nutrient.plots)}
NPlot <- ggplot(nutrients.skinny) +
  geom_point(aes(x = Date, y = TotalNitrogen_mglN)) +
  labs(y = "Total Nitrogen (mg/L)")

print(NPlot)

PPlot <- ggplot(nutrients.skinny) +
  geom_point(aes(x = Date, y = TotalPhosphorus_mglP)) +
  labs(y = "Total Phosphorus (mg/L)")

print(PPlot)

Nutrients.plot <- plot_grid(NPlot,PPlot, ncol = 1) 

print(Nutrients.plot)

```

9. What do these plots tell you about nutrient concentrations over time? How might this relate to your previous plot of hypoxia events?

> Nutrient concentrations were high in the 1980s, and then decreased into the 1990s and 2000s. 

## Discharge and Temperature

10. Turbulent flow in rivers mixes oxygo[en into the water column. As discharge decreases, water moves slower, and oxygen diffuses slower into the water from the atmosphere. Download and reformat the daily discharge data for New Hope Creek (function `readNWISdv()`, site `02097314`, parameter `00060`). 

```{r bring_discharge}
discharge.dat <- readNWISdv(siteNumbers = "02097314", #New Hope Creek
                     parameterCd = "00060", #daily discharge data
                     startDate = "",
                     endDate = "")
names(discharge.dat)[4:5] <- c("Discharge", "Approval.Code")

discharge.skinny <- discharge.dat %>%
  select(Date,Discharge)

```

11. Create a dygraph of discharge and oxygen concentrations. You should have discharge on the y axis and oxygen concentration on the y2 axis. Set your y-axes limits so that they don't change as you zoom in and out. (hint: use `dyAxis(..., valueRange = c(0, 6500))` for discharge & `dyAxis(..., valueRange = c(0, 13))` for oxygen).
```{r}
Oxygen <- with(dat.02, xts(x = Oxygen, order.by = Date))
Discharge <- with(discharge.skinny, xts(x = Discharge, order.by = Date))

DyDat <- cbind(Oxygen, Discharge)

dygraph(DyDat) %>% 
  dySeries("Oxygen", axis = "y2") %>% 
  dyAxis(name = "y", label = "Discharge (cfs)") %>%
  dyAxis(name = "y2", label = "Oxygen (mg/L)", valueRange = c(0, 13)) %>%
  dyRangeSelector()

```

12. Do hypoxic events happen in the wet or the dry season? How many days after storm events (of at least 100 cfs) do they occur? 

> Hypoxic events happen during the dry season. Looking at 1983, the last large discharge event occurred May 28, and the measured hypoxic level was on August 8, so approximately 41 days. When looking at 1984, a rain event occurred on August 12th, followed by a hypoxic measurement on August 22nd, so this only took 10 days. However, the rain event in 1983 had a larger cfs than the one in 1984, which may explain the longer time period before hypoxic conditions set in. 

13. Temperature also impacts the solubility of oxygen in water. More oxygen can dissolve when temperatures are *lower*. How would you analyze the effects of temperature on dissolved oxygen concentrations? What data would you need? How would you separate the effects of temperature from the effects of flow?

> I would need temperature data and dissolved oxygen data. I graph dissolved oxygen and temperature on the same plot and compare the trends to each other. I would also observe flow data and see if that had any significant trends that matched up with temperature and/or dissolved oxygen, potentially excluding measurements that had high cfs on that date.

14. What do you think is the biggest driver of hypoxia in New Hope Creek? What data do you use to support your conclusion?

> I believe the biggest driver of hypoxia in New Hope Creek is from nutrient loading and algal blooms because low flow rate does not seem to be the driver of dissolved oxygen levels after the 1980s. Therefore, hypoxia must be driven by a different factor than flow.

## Reflection
15. What are 2-3 conclusions or summary points about river water quality you learned through your analysis?

> Dissolved oxygen is not significantly impacted by discharge in New Hope Creek, and that Nitrogen and Phosphorus have a greater impact on dissolved oxygen levels in rivers than discharge.

16. What data, visualizations, and/or models supported your conclusions from 15?

> The peaks for Nitrogen and Phosphorus levels coincided with violations of dissolved oxygen levels, and after nutrient levels were consistently low, dissolved oxygen did not have any more violations. 

17. Did hands-on data analysis impact your learning about water quality relative to a theory-based lesson? If so, how?

>Being able to see discharge and dissolved oxygen together helped my understand the weak relationship between the two.

18.	How did the real-world data compare with your expectations from theory?

> I would have expected discharge to have a greater impact on dissolved oxygen. This could possibly have something to do with the speed of flow or the slope of the riverbed.
