---
title: "Assignment 6: Time Series Analysis"
author: "Lindsay Roth"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Hydrologic Data Analysis on time series analysis

## Directions
1. Change "Student Name" on line 3 (above) with your name.
2. Work through the steps, **creating code and output** that fulfill each instruction.
3. Be sure to **answer the questions** in this assignment document.
4. When you have completed the assignment, **Knit** the text and code into a single pdf file.
5. After Knitting, submit the completed exercise (pdf file) to the dropbox in Sakai. Add your last name into the file name (e.g., "A06_Salk.html") prior to submission.

The completed exercise is due on 11 October 2019 at 9:00 am.

## Setup

1. Verify your working directory is set to the R project file, 
2. Load the tidyverse, lubridate, trend, and dataRetrieval packages. 
3. Set your ggplot theme (can be theme_classic or something else)
4. Load the ClearCreekDischarge.Monthly.csv file from the processed data folder. Call this data frame ClearCreekDischarge.Monthly.

```{r setup}
getwd()

library(tidyverse)
library(lubridate)
library(trend)
library(dataRetrieval)

theme_set(theme_classic())

ClearCreekDischarge.Monthly <- read.csv("./Data/Processed/ClearCreekDischarge.Monthly.csv")

```

## Time Series Decomposition
5. Create a new data frame that includes daily mean discharge at the Eno River for all available dates (`siteNumbers = "02085070"`). Rename the columns accordingly.
6. Plot discharge over time with geom_line. Make sure axis labels are formatted appropriately.
7. Create a time series of discharge
8. Decompose the time series using the `stl` function.
9. Visualize the decomposed time series.
```{r}
EnoDischarge <- readNWISdv(siteNumbers = "02085070",
                     parameterCd = "00060", # discharge (ft3/s)
                     startDate = "",
                     endDate = "")

names(EnoDischarge)[4:5] <- c("Discharge", "Approval.Code")

EnoDischargePlot <- ggplot(EnoDischarge) +
  geom_line(aes(x = Date, y = Discharge)) +
  labs(x = "Measurement Dates 1963-2019", y = expression("Discharge (ft"^3*"/s)"))
print(EnoDischargePlot)


EnoDischarge_ts <- ts(EnoDischarge[[4]], frequency = 365)

Eno_Decomposed <- stl(EnoDischarge_ts, s.window = "periodic")

plot(Eno_Decomposed)


ClearCreekDischarge_ts <- ts(ClearCreekDischarge.Monthly[[3]], frequency = 12)

ClearCreek_Decomposed <- stl(ClearCreekDischarge_ts, s.window = "periodic")

plot(ClearCreek_Decomposed)


```

10. How do the seasonal and trend components of the decomposition compare to the Clear Creek discharge dataset? Are they similar in magnitude?

> Seasonal: The seasonal component has a higher magnitude for Clear Creek has a higher magnitude than the Eno. This makes sense because Clear Creek is monthly data and the Eno discharge data is daily.

> Trend: The trend component has approximately equal magnitude for both Clear Creek and Eno, with Clear Creek's being slightly larger.  Both of them have irregular patterns with no clear directional change.

## Trend Analysis
Research question: Has there been a monotonic trend in discharge in Clear Creek over the period of study?

11. Generate a time series of monthly discharge in Clear Creek from the ClearCreekDischarge.Monthly data frame. This time series should include just one column (discharge). 
12. Run a Seasonal Mann-Kendall test on the monthly discharge data. Inspect the overall trend and the monthly trends. 
```{r}
ClearCreek_ts <- ts(ClearCreekDischarge.Monthly[[3]], frequency = 12)

# Run SMK test
ClearCreektrend <- smk.test(ClearCreek_ts)

# Inspect results
ClearCreektrend
summary(ClearCreektrend)

```

13. Is there an overall monotonic trend in discharge over time? If so, is it positive or negative?

> Because the p-value is not less than 0.05, there is not a significant overall monotonic trend in discharge over time.  If the p-value would have been below 0.05, the trend would have been positive since the z-score was greater than 0.

14. Are there any monthly monotonic trends in discharge over time? If so, during which months do they occur and are they positive or negative?

> There were two monthly monotonic trends in discharge over time for the months of June and July (p = 0.047 and p = 0.025, respectively). These trends were both positive because they poth had z-scores greater than 0.

## Reflection
15. What are 2-3 conclusions or summary points about time series you learned through your analysis?

>

16. What data, visualizations, and/or models supported your conclusions from 12?

> 

17. Did hands-on data analysis impact your learning about time series relative to a theory-based lesson? If so, how?

>

18.	How did the real-world data compare with your expectations from theory?

>
